---
title: "Clients, requests and the rest of HTTP"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`zu_get()` and `zu_post()` cover most scripts. This article covers what
comes next: the other methods and body types, reusable clients that hold a
base URL and shared settings, requests you build first and send later, and
the connection pool behind all of them.

Every request here goes to a local test server — an
[httpbin](https://httpbin.org)-style app from the
[webfakes](https://webfakes.r-lib.org) package, running in a background R
process — so the output is real zuhttp output, and the article builds with no
internet access.

```{r server}
library(zuhttp)

srv <- webfakes::new_app_process(
  webfakes::httpbin_app(),
  opts = webfakes::server_opts(remote = TRUE, enable_keep_alive = TRUE,
                               num_threads = 4)
)
base <- srv$url()
base
```

httpbin's endpoints echo the request back as JSON, which makes it easy to see
exactly what was sent. A small helper keeps that readable:

```{r echo-helper}
sent <- function(r) zu_resp_json(r)
```

## The other methods

Every method has a verb, and they share one argument list:

```{r verbs}
r <- zu_put(paste0(base, "put"), json = list(id = 7, name = "Ada"))
sent(r)$json

r <- zu_patch(paste0(base, "patch"), json = list(name = "Ada L."))
sent(r)$json

zu_resp_status(zu_delete(paste0(base, "delete")))
```

`zu_head()` asks for the headers alone — useful to check a resource's type
or freshness without downloading it. The response has headers and no body:

```{r head}
r <- zu_head(paste0(base, "html"))
zu_resp_status(r)
zu_resp_header(r, "content-type")
length(zu_resp_raw(r))
```

## Bodies

Four body arguments, mutually exclusive. The semantic ones set the
`Content-Type` for you, unless you set it yourself.

```{r bodies}
# A form, as a browser would send it.
r <- zu_post(paste0(base, "post"), form = list(user = "ada", lang = "R"))
sent(r)$form

# Raw bytes or a string, with the type you choose.
r <- zu_post(paste0(base, "post"), body = "plain text body",
             headers = c("Content-Type" = "text/plain"))
sent(r)$data

# A file, streamed from disk.
f <- tempfile(fileext = ".csv")
write.csv(head(mtcars, 3), f, row.names = FALSE)
r <- zu_post(paste0(base, "post"), file = f)
sent(r)$headers[["Content-Type"]]
nchar(sent(r)$data)
```

## Reusable clients

A client holds what a group of requests shares: a base URL, default headers
and query parameters, and policy such as timeouts and retries. Pass it with
`client =` — it is always a named argument.

```{r client}
api <- zu_client(
  base_url = base,
  headers  = c(Accept = "application/json", "X-Client" = "article"),
  query    = list(lang = "en"),
  timeout  = 10
)
api
```

Paths are appended to the base URL. It is a join, not RFC 3986 resolution, so
a base URL ending in `/v1` keeps its `/v1`:

```{r client-use}
r <- zu_get("anything/users", query = list(page = 2), client = api)
sent(r)$url
sent(r)$headers[c("Accept", "X-Client")]
```

### How request settings combine with the client's

Headers and query parameters **merge**: the request adds to the client's, and
wins where both set the same name. `NA` removes one the client set.

```{r merge-headers}
r <- zu_get("headers", headers = c(Accept = "text/csv", "X-Client" = NA),
            client = api)
h <- sent(r)$headers
h[["Accept"]]                # the request's value won
"X-Client" %in% names(h)     # removed for this request
```

Policy settings — `timeout`, `retry`, `redirects`, `verify` and the rest —
are **replaced** by a request value. `NULL` resets one to the package default
rather than inheriting the client's.

```{r merge-policy}
zu_resp_status(zu_get("get", timeout = 2, client = api))     # 2 s, this request only
zu_resp_status(zu_get("get", timeout = NULL, client = api))  # the package default, not 10 s
```

### Deriving clients

`zu_client_update()` returns a new client and leaves the original alone —
the usual way to add credentials for one part of an API.

```{r derive}
admin <- zu_client_update(api, headers = c(Authorization = "Bearer admin-token"))
sent(zu_get("headers", client = admin))$headers[["Authorization"]]
"Authorization" %in% names(sent(zu_get("headers", client = api))$headers)
```

If you prefer the client first, pipe-friendly wrappers read that way:

```{r client-first}
api |> zu_client_get("get") |> zu_resp_status()
```

## Build a request now, send it later

Everything the verbs do is available as a pipeline over a request object. No
network I/O happens until `zu_perform()`, so a package can build a request
in one function, inspect or test it, and send it in another.

```{r compose}
req <- zu_request("POST", paste0(base, "anything/orders")) |>
  zu_query(dry_run = "true") |>
  zu_headers("Idempotency-Key" = "order-2026-001") |>
  zu_body_json(list(item = "book", qty = 2)) |>
  zu_req_timeout(total = 5)

req
```

The printed request hides nothing it needs and shows nothing it should not:
credentials would read `<redacted>` (see the
[debugging article](debugging.html)). Then:

```{r perform}
r <- zu_perform(req, client = api)
sent(r)$args      # the query: the request's dry_run, the client's lang
sent(r)$json
```

`zu_get(url, ...)` and `zu_request("GET", url) |> ... |> zu_perform()` lower
to the same request, so they behave identically.

## Connection reuse

A client keeps a pool of open connections, so a second request to the same
host skips the TCP (and TLS) setup. `zu_pool_stats()` shows it happening:

```{r pool}
pooled <- zu_client(base_url = base)
for (i in 1:5) zu_get("get", client = pooled)
zu_pool_stats(pooled)[c("hits", "misses", "idle")]
```

One miss opens the connection; the next four reuse it. A connection is only
reused when every setting that affects it matches — host, port, proxy and
the whole TLS configuration — so a request with different trust settings
never rides on a connection made without them. The pool's limits are set
with `zu_pool()`, and `zu_client(pool = NULL)` turns pooling off.

The pool is also safe across `fork()`: a child process never writes into a
connection its parent owns. For parallel work, use a PSOCK cluster or
`future::plan("multisession")`; on macOS, HTTPS in a forked child raises
`zu_fork_error` rather than crashing (`?zuhttp_fork`).

## The default client

The one-shot calls use a package-level default client. Replace it to change
defaults for a whole session; `zu_info()` reports what it is set to.

```{r default-client}
old <- zu_set_default_client(zu_client(timeout = 15, headers = c("X-Team" = "data")))
sent(zu_get(paste0(base, "headers")))$headers[["X-Team"]]
zu_info()$default_client$timeout
zu_set_default_client(old)   # restore
```

```{r cleanup, include = FALSE}
srv$stop()
```
