---
title: "Seeing what happened: verbose output, traces, hooks and middleware"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

When a request misbehaves, the first question is what actually happened on
the wire. zuhttp answers it at several levels of detail: a printed response,
verbose logging, a full event trace, timings and connection metadata. For
behaviour you want every time, there are hooks, which observe, and
middleware, which can change the request.

All of it goes through one redaction policy, so none of it prints your
credentials.

```{r server}
library(zuhttp)
srv <- webfakes::new_app_process(
  webfakes::httpbin_app(),
  opts = webfakes::server_opts(remote = TRUE, enable_keep_alive = TRUE,
                               num_threads = 4)
)
url <- function(path) srv$url(path)
```

## Printing is safe

A printed request or response shows what you need to recognise it — method,
URL, headers, body size, timing — with secret values replaced:

```{r print}
req <- zu_request("GET", url("/get")) |>
  zu_query(api_key = "KEY-12345", page = 2) |>
  zu_headers(Authorization = "Bearer s3cret", Accept = "application/json")
req
```

The request itself is untouched: redaction happens when formatting, so the
request above still sends the real key and token. The same applies to error
messages and condition objects, which routinely end up in logs and bug
reports.

## Verbose output

`zu_verbose()` narrates each request as it happens, on standard error so it
never mixes with data a script writes to standard output. Add `trace = TRUE`
to see the phases underneath — DNS, connect, request, headers, body — with
timestamps:

```{r verbose}
chatty <- zu_client(hooks = zu_verbose(to = stdout()))
r <- zu_get(url("/gzip"), trace = TRUE, client = chatty)
```

(`to = stdout()` is only so the output appears in this article.)

## The event trace

With `trace = TRUE` the same events are kept on the response, as a data
frame you can inspect or save:

```{r trace}
r <- zu_get(url("/redirect/1"), trace = TRUE)
zu_resp_trace(r)
```

A followed redirect appears as an event carrying the resolved target, and a
URL is redacted as it is recorded.

## Timings and connection details

Every response carries its timings, in seconds. A phase that did not happen
is `NA`, not zero — a reused connection has no DNS or connect time:

```{r timings}
api <- zu_client(base_url = srv$url())
first  <- zu_get("get", client = api)
second <- zu_get("get", client = api)
rbind(first = zu_resp_timings(first), second = zu_resp_timings(second))[
  , c("dns", "connect", "ttfb", "total")]
```

`zu_resp_connection()` answers the questions a timing cannot:

```{r connection}
str(zu_resp_connection(first))
zu_resp_connection(second)$reused_connection
```

Over HTTPS it also reports the negotiated protocol and cipher, and which
trust store vouched for the certificate.

## What this build can do

`zu_info()` is the thing to paste into a bug report: the TLS backend and trust
store, the `zu_tls()` settings this platform supports, the default client, and
which proxy variables are set (with any credentials in them redacted):

```{r info}
zu_info()
```

## Hooks: observe every request

Hooks are functions zuhttp calls at points in a request's life:
`before_request`, `after_response`, `before_retry` and `after_retry`. They
see a redacted copy of the request and response, and their return value is
ignored — a hook can log, count or time, but cannot change what is sent. A
hook that errors is downgraded to a warning, so observability never breaks a
request.

A simple request log:

```{r hooks}
log <- data.frame()
logging <- zu_client(
  base_url = srv$url(),
  hooks = zu_hooks(after_response = function(p) {
    log <<- rbind(log, data.frame(
      method = p$request$method,
      url    = p$request$url,
      status = zu_resp_status(p$response),
      ms     = round(1000 * zu_resp_timings(p$response)[["total"]], 1)
    ))
  })
)
invisible(zu_get("get", client = logging))
invisible(zu_post("post", json = list(a = 1), client = logging))
invisible(zu_get("status/404", check = FALSE, client = logging))
log
```

## Middleware: change requests and responses

Middleware is a function of the request and the next step, returning a
response. It wraps everything below it — including retries, so a middleware
sees one logical request however many attempts it takes. The first in the
list is the outermost.

Adding a header to every request, such as a signature or a request ID:

```{r middleware-header}
add_request_id <- function(req, next_fn) {
  req$headers <- c(req$headers, "X-Request-Id" = sprintf("req-%04d", sample(9999, 1)))
  next_fn(req)
}
signed <- zu_client(base_url = srv$url(), middleware = list(add_request_id))
zu_resp_json(zu_get("headers", client = signed))$headers[["X-Request-Id"]]
```

Middleware can also answer without calling `next_fn` at all. A minimal
in-memory cache:

```{r middleware-cache}
cache <- new.env()
cached <- function(req, next_fn) {
  key <- paste(req$method, req$url)
  if (req$method == "GET" && !is.null(cache[[key]])) return(cache[[key]])
  resp <- next_fn(req)
  if (req$method == "GET" && zu_resp_ok(resp)) cache[[key]] <- resp
  resp
}
memo <- zu_client(base_url = srv$url(), middleware = list(cached))
a <- zu_get("uuid", client = memo)
b <- zu_get("uuid", client = memo)   # served from the cache
identical(zu_resp_json(a)$uuid, zu_resp_json(b)$uuid)
```

Use a hook when you only need to watch, and middleware when you need to act.
Policies zuhttp already has — timeouts, retries, redirects, TLS — are set
with their own arguments, not rebuilt as middleware.

## Redaction, in detail

The same rules apply everywhere output leaves zuhttp: printing, error
messages, hooks, verbose output, traces, cassettes and `zu_resp_url()`.

- **Headers:** `Authorization`, `Proxy-Authorization`, `Cookie`,
  `Set-Cookie`, `X-Api-Key` and `X-Auth-Token`.
- **URLs:** any `user:password@` part, and the values of query parameters
  with names like `access_token`, `api_key`, `token`, `password` or
  `client_secret` (matched exactly, so `page_token` is left alone).
- **Form bodies:** the same parameter names.

```{r redact}
zu_redact_url("https://user:pw@api.example.com/v1?access_token=abc&page_token=2")
zu_is_secret_param(c("token", "page_token"))
```

Your API may use its own names. Add them for the session — they extend the
defaults and cannot remove one:

```{r redact-custom}
old <- getOption("zuhttp.redact_params")
zu_redact_params("ticket")
zu_redact_url("https://api.example.com/v1?ticket=T-998877")
options(zuhttp.redact_params = old)
```

```{r cleanup, include = FALSE}
srv$stop()
```
