---
title: "When requests fail: errors, timeouts, retries and redirects"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Requests fail in two different ways, and zuhttp keeps them apart. A
**transport failure** means no usable response arrived: the host did not
resolve, the connection was refused, the certificate was wrong, the clock ran
out. An **HTTP error** means the server answered, and the answer was no. Both
are R conditions with classes you can catch, and neither is a string you have
to parse.

This article runs against two local test servers
([webfakes](https://webfakes.r-lib.org)). One gets two extra endpoints that
fail on purpose, so retries have something to retry:

```{r servers}
library(zuhttp)

app <- webfakes::httpbin_app()
# Handlers run in the server process, so each keeps its own count there.
app$locals$flaky <- 0L
app$locals$orders <- 0L
# Fails twice with "503, try again in 1 second", then succeeds.
app$get("/flaky", function(req, res) {
  n <- req$app$locals$flaky <- req$app$locals$flaky + 1L
  if (n %% 3L != 0L) {
    res$set_status(503L)$set_header("Retry-After", "1")$send("busy")
  } else {
    res$send_json(list(ok = TRUE, attempt = n), auto_unbox = TRUE)
  }
})
# The same, for a POST.
app$post("/flaky-order", function(req, res) {
  n <- req$app$locals$orders <- req$app$locals$orders + 1L
  if (n %% 3L != 0L) res$set_status(503L)$send("busy")
  else res$send_json(list(created = TRUE, attempt = n), auto_unbox = TRUE)
})

srv   <- webfakes::new_app_process(app)
other <- webfakes::new_app_process(webfakes::httpbin_app())  # another origin
url <- function(path) srv$url(path)
```

## HTTP errors are conditions

By default a 4xx or 5xx response raises. The message says what was attempted,
what came back, and what usually fixes it:

```{r status-error, error = TRUE}
zu_get(url("/status/404"))
```

Every zuhttp condition inherits from `zu_error`, and the classes form a tree,
so you catch as broadly or as narrowly as you need:

```{r classes}
e <- tryCatch(zu_get(url("/status/503")), error = function(e) e)
class(e)

tryCatch(zu_get(url("/status/404")),
         zu_http_client_error = function(e) "a 4xx: fix the request",
         zu_http_server_error = function(e) "a 5xx: maybe try again")
```

The condition carries structured fields as well as a message, including the
response itself when there is one:

```{r fields}
zu_resp_status(e$response)
zu_resp_header(e$response, "content-type")
```

To treat an error status as data instead, turn the check off, per request or
for a whole client, and decide later with `zu_resp_check()`:

```{r check-false}
r <- zu_get(url("/status/404"), check = FALSE)
zu_resp_status(r)
zu_resp_ok(r)
tryCatch(zu_resp_check(r), zu_http_status_error = function(e) "checked later")
```

## Transport failures

The same tree covers failures below HTTP. Each class names the phase that
failed, which is the first thing to know when debugging one:

```{r transport, error = TRUE}
# Nothing listens on this port any more.
closed <- webfakes::new_app_process(webfakes::httpbin_app())
closed_url <- closed$url("/get")
closed$stop()
e <- tryCatch(zu_get(closed_url), error = function(e) e)
class(e)[1]
e$phase
e$retryable   # a failed connection may well succeed next time

# Not a URL zuhttp will send.
tryCatch(zu_get("ftp://example.com/file"), zu_url_error = function(e) conditionMessage(e))
```

## Timeouts

`timeout` is a budget in seconds for the whole call: every redirect, every
retry, every backoff between them. When it runs out, `zu_timeout_error` is
raised, however far the request got:

```{r timeout}
t0 <- Sys.time()
e <- tryCatch(zu_get(url("/delay/5"), timeout = 1), error = function(e) e)
class(e)[1]
round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1)
```

The request stopped after about a second, not five. A slow download that
keeps making progress still counts against the same budget, so set `timeout`
with the largest response you expect in mind. There are no separate connect
or inactivity timeouts yet, and name resolution is not bounded by `timeout`
(see `?zuhttp_tls`).

## Retries

Retrying is **off by default**: a client that silently repeats requests can
repeat a payment. Turn it on with a policy:

```{r retry}
r <- zu_get(url("/flaky"), retry = zu_retry(attempts = 3))
zu_resp_json(r)
zu_resp_connection(r)$retries_performed
```

`attempts` counts the first try, so `attempts = 3` means at most two retries.
Between attempts zuhttp waits: exponential backoff with jitter, or what the
server asks for in `Retry-After` — here, one second — capped by
`max_retry_after` so a server cannot park your session for an hour. A hook
shows each decision as it happens:

```{r retry-hook}
watch <- zu_client(
  retry = zu_retry(attempts = 3),
  hooks = zu_hooks(before_retry = function(ctx) {
    message("retry ", ctx$attempt, " in ", ctx$delay, "s after ", ctx$why)
  })
)
r <- zu_get(url("/flaky"), client = watch)
zu_resp_status(r)
```

Only failures that might succeed next time are retried: connection failures,
timeouts with budget left, and the statuses 408, 429, 500, 502, 503 and 504.
A certificate error or a 404 is never retried.

### A POST is not retried unless you say it is safe

`GET`, `PUT` and `DELETE` are idempotent: sending one twice has the same
effect as sending it once. `POST` is not, so a retry policy does not apply to
it, and the first 503 is the answer:

```{r post-not-retried}
e <- tryCatch(
  zu_post(url("/flaky-order"), json = list(item = "book"),
          retry = zu_retry(attempts = 3)),
  error = function(e) e
)
class(e)[1]
```

When the server makes a POST safe to repeat — most payment and order APIs do,
through an `Idempotency-Key` header — say so on the request.
`zu_req_retry()` generates the key and marks the request replay-safe:

```{r post-retried}
req <- zu_request("POST", url("/flaky-order")) |>
  zu_body_json(list(item = "book")) |>
  zu_req_retry(attempts = 3, idempotency_key = TRUE)
zu_req_replay_safe(req)
zu_resp_json(zu_perform(req))
```

## Redirects

zuhttp follows redirects, up to 10 by default, and counts them:

```{r redirects}
r <- zu_get(url("/redirect/3"))
zu_resp_url(r)
zu_resp_connection(r)$redirect_count
```

A chain longer than `redirects` is an error rather than a surprise 3xx.
`redirects = 0` asks for the redirect itself:

```{r redirect-limit, error = TRUE}
zu_get(url("/redirect/3"), redirects = 2)

r <- zu_get(url("/redirect/1"), redirects = 0)
zu_resp_status(r)
zu_resp_header(r, "location")
```

A 303 turns any method into a `GET` and drops the body; 307 and 308 keep
both. That is why a body you want replayed must be rewindable — memory and
files are.

```{r redirect-methods}
see_other <- paste0(url("/redirect-to"), "?status_code=303&url=",
                    utils::URLencode(url("/anything"), reserved = TRUE))
zu_resp_json(zu_post(see_other, body = "order"))$method
```

### Credentials stay with their origin

A redirect to a different scheme, host or port drops `Authorization`,
`Cookie` and `Proxy-Authorization`, so a token meant for one server is never
handed to another. Other headers travel on:

```{r strip}
to_other <- paste0(url("/redirect-to"), "?url=",
                   utils::URLencode(other$url("/headers"), reserved = TRUE))
r <- zu_get(to_other, headers = c(Authorization = "Bearer s3cret",
                                  "X-Request-Id" = "abc-123"))
sent <- zu_resp_json(r)$headers
sent[["X-Request-Id"]]
is.null(sent$Authorization)
```

A redirect from `https://` to `http://` is refused outright with
`zu_redirect_error`.

```{r cleanup, include = FALSE}
srv$stop()
other$stop()
```
