---
title: "Large responses: downloads, streaming and limits"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

By default a response body is read into memory, which is right for an API
reply and wrong for a 2 GB file. zuhttp has two other destinations — a file,
and a function you supply — and in both the body is never held in memory
whole. This article also covers compression, and the limits that stop a
response from growing without bound.

The examples use a local test server ([webfakes](https://webfakes.r-lib.org))
with one extra endpoint, a compressed response that would inflate to 50 MB:

```{r server}
library(zuhttp)

app <- webfakes::httpbin_app()
app$get("/bomb", function(req, res) {
  # 50 MB of zeros, gzip-compressed on the fly: about 50 KB on the wire.
  tf <- tempfile()
  con <- gzfile(tf, "wb"); writeBin(raw(50e6), con); close(con)
  res$set_header("Content-Encoding", "gzip")$
    set_type("application/octet-stream")$
    send(readBin(tf, "raw", file.size(tf)))
})
srv <- webfakes::new_app_process(app)
url <- function(path) srv$url(path)
```

## Downloading to a file

`path =` writes the body straight to disk:

```{r path}
dest <- tempfile(fileext = ".bin")
r <- zu_get(url("/bytes/10000"), path = dest)
zu_resp_path(r)
file.size(dest)
length(zu_resp_raw(r))   # nothing was kept in memory
```

The file appears at `dest` only once the body has arrived whole. zuhttp
writes to a temporary file beside it and renames at the end, so a download
that fails part-way leaves nothing behind, and a file that was already at
`dest` is untouched:

```{r path-atomic}
writeLines("the previous version", dest)

# Five bytes spread over three seconds, and a one-second budget.
e <- tryCatch(
  zu_get(url("/drip?numbytes=5&duration=3"), path = dest, timeout = 1),
  error = function(e) e
)
class(e)[1]
readLines(dest)            # the old file is still intact
list.files(dirname(dest), pattern = basename(dest))   # and no partial file
```

One thing a file sink does not do is judge the status code. A 404 page is a
complete body, so with `check = FALSE` it is written like any other; check
`zu_resp_ok()` before trusting the file.

## Streaming to a function

`callback =` receives the body in chunks as they arrive, already
decompressed. Return `FALSE` to stop. Here the server sends five bytes over
two seconds, and the callback sees them as they come:

```{r callback}
seen <- character()
r <- zu_get(url("/drip?numbytes=5&duration=2"), callback = function(chunk) {
  seen <<- c(seen, format(Sys.time(), "%H:%M:%OS1"))
  TRUE
})
length(seen)      # one call per chunk
seen
```

That makes line-oriented streams easy to process incrementally — for example
newline-delimited JSON, parsed a record at a time without ever holding the
whole response:

```{r ndjson}
records <- list()
pending <- ""
r <- zu_get(url("/stream/5"), callback = function(chunk) {
  lines <- strsplit(paste0(pending, rawToChar(chunk)), "\n", fixed = TRUE)[[1]]
  pending <<- if (endsWith(rawToChar(chunk), "\n")) "" else lines[length(lines)]
  complete <- if (nzchar(pending)) lines[-length(lines)] else lines
  for (l in complete[nzchar(complete)]) {
    records[[length(records) + 1]] <<- jsonlite::fromJSON(l)$id
  }
  TRUE
})
unlist(records)
```

Returning `FALSE` ends the transfer cleanly — useful when the first part of a
response already answers the question:

```{r stop-early}
chunks <- 0
r <- zu_get(url("/drip?numbytes=10&duration=2"), callback = function(chunk) {
  chunks <<- chunks + 1
  chunks < 2        # stop after the second chunk
})
chunks
zu_resp_status(r)   # the response is still a response
```

An error inside the callback stops the transfer too, and reaches you as the
error you raised, not wrapped in a zuhttp one:

```{r callback-error, error = TRUE}
zu_get(url("/stream/5"), callback = function(chunk) stop("not what I expected"))
```

## Compression

zuhttp asks for gzip and decompresses transparently. `zu_resp_raw()` and
`zu_resp_text()` return the decoded body, and the timings show what actually
crossed the wire:

```{r gzip}
r <- zu_get(url("/gzip"))
zu_resp_json(r)$gzipped
zu_resp_timings(r)[c("body_bytes_wire", "body_bytes_decoded")]
```

`decode = FALSE` keeps the bytes exactly as sent — what you want when
checking them against a published checksum:

```{r no-decode}
r <- zu_get(url("/gzip"), decode = FALSE)
zu_resp_header(r, "content-encoding")
head(zu_resp_raw(r), 4)                 # the gzip magic number, 1f 8b
```

## Limits

A response body is capped at 16 MiB by default, counted *after*
decompression, on every destination — memory, file or callback. Raise it with
`max_body` when you expect more:

```{r max-body, error = TRUE}
zu_get(url("/bytes/5000"), max_body = 1000)
```

The limit is enforced while decompressing, which is what makes it a defence
and not only a setting. The `/bomb` endpoint sends about 50 KB that would
inflate to 50 MB; zuhttp stops at the limit instead of allocating the rest:

```{r bomb, error = TRUE}
zu_get(url("/bomb"))
```

For a file you expect to be large, raise `max_body` on that request, or on a
client used for downloads.

```{r cleanup, include = FALSE}
srv$stop()
unlink(dest)
```
