---
title: "Testing code that makes HTTP requests"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Code that talks to a web API is awkward to test: the service may be slow,
rate-limited, need credentials, or be down on the day CRAN checks your
package. zuhttp lets you replace the network at one point — the client's
**transport** — while everything above it (URL building, headers, retries,
response parsing, error handling) runs for real.

There are three levels, from simplest to most realistic:

| Tool | What it does | Use it for |
|---|---|---|
| `zu_mock_transport(function)` | calls your function for every request | full control, computed replies |
| `zu_stub()` | matches requests and returns canned responses | most unit tests |
| `zu_cassette_transport()` | records real traffic once, replays it after | tests shaped like the real API |

```{r setup}
library(zuhttp)
```

## Write functions that accept a client

The one design choice that makes all of this work: functions in your package
take a `client` argument, defaulting to a real one. Tests pass a different
client; users never notice.

```{r package-code}
# In your package:
github_client <- function() {
  zu_client(base_url = "https://api.github.com",
            headers = c(Accept = "application/vnd.github+json"))
}

repo_stars <- function(repo, client = github_client()) {
  r <- zu_get(paste0("repos/", repo), client = client)
  zu_resp_json(r)$stargazers_count
}
```

## A mock transport

`zu_mock_transport()` takes a function of the request and returns a response
built with `zu_response()`. No network is involved:

```{r mock}
fake_github <- zu_mock_transport(function(req) {
  zu_response(200L, c("Content-Type" = "application/json"),
              '{"full_name": "r-lib/httr2", "stargazers_count": 250}')
})
test_client <- zu_client_update(github_client(), transport = fake_github)

repo_stars("r-lib/httr2", client = test_client)
```

The function sees the request as zuhttp would send it — after the client's
base URL, headers and query have been merged — so it can assert on it too:

```{r mock-inspect}
seen <- NULL
spy <- zu_mock_transport(function(req) {
  seen <<- req
  zu_response(200L, body = '{"stargazers_count": 1}')
})
invisible(repo_stars("r-lib/httr2",
                     client = zu_client_update(github_client(), transport = spy)))
seen$method
seen$url
```

## Stubs: canned responses by pattern

For most tests, list the requests you expect and what each should get back.
A stub matches on method, URL (exact or by `regex`), headers and body; every
criterion you give must match.

```{r stubs}
api <- zu_client(
  base_url = "https://api.example.com",
  transport = zu_mock_transport(
    zu_stub(zu_response(200L, body = '[{"id": 1}, {"id": 2}]'),
            method = "GET", regex = "/users$"),
    zu_stub(zu_response(201L, body = '{"id": 3}'),
            method = "POST", regex = "/users$"),
    zu_stub(zu_response(404L, body = "no such user"),
            regex = "/users/[0-9]+$")
  )
)

zu_resp_json(zu_get("users", client = api))$id
zu_resp_status(zu_post("users", json = list(name = "Ada"), client = api))
tryCatch(zu_get("users/99", client = api),
         zu_http_client_error = function(e) "404, as expected")
```

A request that matches no stub is an error, so a test cannot pass by
accidentally hitting an endpoint you did not plan for:

```{r unmatched, error = TRUE}
zu_get("orders", client = api)
```

### Testing failure paths

The cases that are hard to provoke against a real server are the ones a mock
makes easy. `times` limits how often a stub may match, so a sequence of stubs
describes a server that fails and then recovers. Here, a 503 followed by a
success tests that retries are wired up:

```{r stub-retry}
flaky <- zu_client(
  transport = zu_mock_transport(
    zu_stub(zu_response(503L), times = 1),
    zu_stub(zu_response(200L, body = "ok"))
  ),
  retry = zu_retry(attempts = 2, base = 0.01)
)
r <- zu_get("https://api.example.com/report", client = flaky)
zu_resp_text(r)
zu_resp_connection(r)$retries_performed
```

And `times = 1` with no fallback asserts the opposite — that a request is
*not* retried. A POST must not be, and if it were, the second attempt would
find no stub:

```{r stub-no-retry}
once <- zu_client(
  transport = zu_mock_transport(zu_stub(zu_response(503L), times = 1)),
  retry = zu_retry(attempts = 3, base = 0.01)
)
e <- tryCatch(zu_post("https://api.example.com/charge", body = "x", client = once),
              error = function(e) e)
class(e)[1]   # the 503 itself: no second attempt was made
```

Transport failures can be simulated by raising the condition zuhttp would
raise, with `zu_condition()`:

```{r mock-timeout}
slow <- zu_client(transport = zu_mock_transport(function(req) {
  stop(zu_condition("zu_timeout_error", "simulated: no response in time"))
}))
tryCatch(repo_stars("r-lib/httr2", client = zu_client_update(github_client(),
                                                             transport = slow$transport)),
         zu_timeout_error = function(e) "handled the timeout")
```

## Cassettes: record once, replay forever

A cassette records real responses the first time a request is made and
replays them afterwards, so tests are written against the real service but
run offline. To keep this article offline too, the "real service" below is a
local test server ([webfakes](https://webfakes.r-lib.org)).

```{r cassette-record}
srv <- webfakes::new_app_process(webfakes::httpbin_app())
dir <- file.path(tempdir(), "cassettes")

json_url <- srv$url("/json")

recorder <- zu_client(transport = zu_cassette_transport(dir, "httpbin"))
r <- zu_get(json_url, client = recorder)
zu_resp_json(r)$firstName
```

Now stop the server. In `"replay"` mode the cassette never touches the
network, and a request it has not seen is an error rather than a live call:

```{r cassette-replay}
srv$stop()
player <- zu_client(transport = zu_cassette_transport(dir, "httpbin", mode = "replay"))
zu_resp_json(zu_get(json_url, client = player))$firstName
```

In a package, point `dir` at `tests/testthat/cassettes/`, record once
(`mode = "auto"`, the default, records only what is missing), commit the
files, and CI replays them.

### Cassettes never contain your credentials

A cassette is a file that ends up in version control, so zuhttp removes
credentials before anything is written: secret headers in both directions,
userinfo in URLs, and secret query and form parameters, all by the same rules
used for printing and error messages (see the
[debugging article](debugging.html)).

```{r cassette-redaction}
srv <- webfakes::new_app_process(webfakes::httpbin_app())
rec <- zu_client(transport = zu_cassette_transport(dir, "secrets"),
                 headers = c(Authorization = "Bearer SUPER-SECRET-TOKEN"))
invisible(zu_get(srv$url("/json"), query = list(api_key = "KEY-12345", page = 2),
                 client = rec))
srv$stop()

stored <- zu_cassette_interactions(dir, "secrets")[[1]]
stored$request$url
stored$request$headers[["Authorization"]]

# And not anywhere in the file's bytes (cassettes are stored uncompressed,
# so a byte search is a real check):
bytes <- readBin(file.path(dir, "secrets.rds"), "raw", 1e6)
length(grepRaw("SUPER-SECRET-TOKEN", bytes, fixed = TRUE))
length(grepRaw("KEY-12345", bytes, fixed = TRUE))
```

Matching uses the redacted request, so rotating a token does not invalidate
a recorded suite.

One thing redaction cannot do is read a **response body** for you. The body
is stored as the server sent it, so an endpoint that echoes your request
(httpbin's `/get` does) or *issues* credentials (an OAuth token endpoint)
puts them in the cassette. Record those endpoints with a mock instead, or
check the cassette before committing it.

```{r cleanup, include = FALSE}
zu_cassette_clear(dir, "httpbin")
zu_cassette_clear(dir, "secrets")
```
